East Coast Sparrow Model Compare Methods to Aggregate Loads to Lakes
========================================================
Data
-------------------------
* East Coast Sparrow data received from Anne Hoos on September 23 2014 as 4 sas files (forBryan_MRB2_Nitrogen_2002vs2020.zip) and saved to C:\Bryan\EPA\Data\Sparrow\EastCoast\Predictions20140923
  * mrb2_massbalancereservoirs_2002.sas7bdat
  * mrb2_massbalancereservoirs_2020.sas7bdat
  * mrb2_predict_tn_2002.sas7bdat
  * mrb2_predict_tn_2020.sas7bdat
* SAS datasets opened in SAS and exported to Access
* These data were added to a new access database (C:\Bryan\EPA\Data\Sparrow\EastCoast\Predictions20140923\EC2014.mdb) and the following tables were added as linked tables
  * MRB1_NHDWaterbody (WaterbodyDatabse.mdb): the MRB1 NHDplus waterbodies showing the COMIDs and Reachcodes.
  * tblJoinCOMID_WBID (WaterbodyDatabse.mdb): a table to match waterbody COMIDs to the Lake WBIDs
  * tblJoinFlowline_WBID_InOut (MRB1.mdb): this table links the flowline COMIDs to the WBIDs and shows whether the flowline is an input or output to the lake or an internal reach.
  * tblWBID_SParrowLoadDSS (MRB1.mdb): table with the Sparrow loads and flow I calculated from the DSS data.
* tblWBRCHCODE was created with the "qryWBRCHCODE_Final".  This table links the WBIDs to the Waterbody Reachcodes.  A more detailed analysis of this relationship is presented below. 

<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
  knit('./ECcompare.rmd')
-->

```{r loadPkgs, include=FALSE, echo=FALSE, cache=FALSE}
  options(stringsAsFactors = FALSE, scipen=9) 


  #function to install (if needed) and load R packages by list
    libs<-c('RODBC','RSQLite') #list of packages to load
  
    installLoad<-function(pck)#user defined function
    {
      if(!pck%in%installed.packages()){install.packages(pck,repos="http://rweb.quant.ku.edu/cran/")}
      require(pck, character.only = TRUE)
    }
  
    lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

```{r gridcodes, include=FALSE, echo=FALSE, cache=TRUE}
#get tblJoinCOMID_WBID & MRB1_NHDWaterbody
  con <- odbcConnectAccess('C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/EC2014.mdb')
    tblJoinCOMID_WBID<-sqlQuery(con,"SELECT tblJoinCOMID_WBID.WB_ID, tblJoinCOMID_WBID.COMID
                                     FROM tblJoinCOMID_WBID;
          ")
      nrow(tblJoinCOMID_WBID)  #37192
      length(unique(tblJoinCOMID_WBID$WB_ID))  #36274
    MRB1_NHDWaterbody<-sqlQuery(con,"SELECT MRB1_NHDWaterbody.COMID, MRB1_NHDWaterbody.REACHCODE
                                    FROM MRB1_NHDWaterbody;
          ")
      nrow(MRB1_NHDWaterbody)  #37190

  close(con)

# RSQLite
  # Create the database if none exists;
  # Or connects to the database if one already exists.
    EC <- dbConnect(SQLite(), dbname='EC.sqlite')
  #use this to delete db
    #if (file.exists("EC.sqlite") == TRUE) file.remove("EC.sqlite") 
  # Import tblJoinCOMID_WBID & MRB1_NHDWaterbody into database
      dbWriteTable(conn=EC,name="tblJoinCOMID_WBID",value=tblJoinCOMID_WBID,row.names=FALSE,overwrite=TRUE)
      dbWriteTable(conn=EC,name="MRB1_NHDWaterbody",value=MRB1_NHDWaterbody,row.names=FALSE,overwrite=TRUE)
        #head(dbReadTable(EC, "tblJoinCOMID_WBID")) 
        #nrow(dbReadTable(EC, "tblJoinCOMID_WBID")) #37192
        #head(dbReadTable(EC, "MRB1_NHDWaterbody")) 
        #nrow(dbReadTable(EC, "MRB1_NHDWaterbody")) #37190
  #join tblJoinCOMID_WBID & MRB1_NHDWaterbody and write new table to database
    a<-dbGetQuery(EC, "
      SELECT tblJoinCOMID_WBID.WB_ID,tblJoinCOMID_WBID.COMID,MRB1_NHDWaterbody.REACHCODE
      FROM tblJoinCOMID_WBID INNER JOIN MRB1_NHDWaterbody ON tblJoinCOMID_WBID.COMID = MRB1_NHDWaterbody.COMID
      GROUP BY tblJoinCOMID_WBID.WB_ID, MRB1_NHDWaterbody.REACHCODE;
    ")
        length(a$WB_ID) #36306
        length(unique(a$WB_ID)) #36274
    dbWriteTable(conn=EC,name="temp",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "temp"))
      #nrow(dbReadTable(EC, "temp"))  36306
  #count the number of gridcodes per WBID
    a<-dbGetQuery(EC, "
      SELECT temp.WB_ID,Count(temp.REACHCODE) AS Count
      FROM temp
      GROUP BY temp.WB_ID
      ORDER BY Count(temp.REACHCODE) DESC
    ")
    dbWriteTable(conn=EC,name="temp1",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "temp1"))
  #add gridcode to temp1 and rename columns
    a<-dbGetQuery(EC, "
      SELECT temp.WB_ID AS WBID,temp.COMID,temp.REACHCODE AS WBRchCd,temp1.Count
      FROM temp INNER JOIN temp1 ON temp.WB_ID=temp1.WB_ID
      ORDER BY temp.WB_ID
    ")
    dbWriteTable(conn=EC,name="tblJoinWBID_WBRchCd",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "tblJoinWBID_WBRchCd"))
  #delete the working tables  
    dbListTables(EC) # List the tables in the database
    dbRemoveTable(EC, "tblJoinCOMID_WBID")
    dbRemoveTable(EC, "MRB1_NHDWaterbody")
    dbRemoveTable(EC, "temp")
    dbRemoveTable(EC, "temp1")
  #read tblJoinWBID_WBRchCd into memory to be used in the text below
      tblJoinWBID_WBRchCd<-dbGetQuery(EC, "
        SELECT tblJoinWBID_WBRchCd.*
        FROM tblJoinWBID_WBRchCd
        ")


head(tblJoinWBID_WBRchCd)
```

Use RSQLite to create tblJoinWBID_WBRchCd 
-------------------------
* First step: join the WBIDs to the GRIDCODES for MRB1
  * Gridcodes for each NHDplus waterbody (COMID) are in the table MRB1_NHDWaterbody in EC2014.mdb
  * The waterbody COMIDs and Gridcodes need to be joined to the WBIDs in tblJoinCOMID_WBID
  * These tables are imported into a new SQLite databse (EC.sqlite)
  * SQL is used to 
    * read the tables from EC.sqlite, join them and write them to a new table.
    * identify the WBIDs with multiple GRIDcodes 
    * create a new table with WBID, COMID, GRIDCODE, and the Count
    * delete the tables temp, temp1, tblJoinCOMID_WBID & MRB1_NHDWaterbody
  * we end with a single table (tblJoinWBID_WBRchCd) in EC.sqlite
  * Evaluate tblJoinWBID_WBRchCd
      * `r nrow(tblJoinWBID_WBRchCd)` observations
      * `r length(unique(tblJoinWBID_WBRchCd$WBID))` unique WBIDs 
      * `r length(unique(tblJoinWBID_WBRchCd$WBRchCd))` unique Reachcodes
      * `r nrow(tblJoinWBID_WBRchCd[is.na(tblJoinWBID_WBRchCd$WBRchCd),])` COMIDs with missing reachcodes; most of these are not really lakes but there is one is a subWaterbody of lake Champlain (WBID=22302965) with a missing reachcode.
      * `r a<-tblJoinWBID_WBRchCd[tblJoinWBID_WBRchCd$Count>1,'WBID'];length(a)` observations with multiple reachcodes representing `r length(unique(a))` WBID lakes
  * Below is a list of the lakes with multiple reachcodes.  These should be removed from the anlaysis.
  
```{r listLakes, include=TRUE, echo=FALSE, cache=FALSE}
  tblJoinWBID_WBRchCd[tblJoinWBID_WBRchCd$Count>1,]
```

Data Definitions for tblJoinWBID_WBRchCd
-------------------------
**Field** | **Definition**
------------- | -------------
**WBID** | Unique Waterbody ID
**COMID** | NHDplus Unique Waterbody ID-sometimes there are multiple COMID for a single WBID
**WBRchCd** | NHDplus Waterbody Reachcode associated with each COMID; Note this was converted to numeric so leading Zero was lost.  Leading Zero can be added back or used as numeric.
**Count** | Number of WBRchCds associated with the WBID-for analysis keep Count==1 

Estimate Nitrogen Loads for Each WBID
-------------------------
**Get the 2002 Nitrogen predictions with the undecayed incremental loads**
* There are a lot of fields most of which will not be used so I am not going to give data definitions
* add table to EC.sqlite 

```{r N2002, include=FALSE, echo=FALSE, cache=TRUE}
#Get the 2002 Nitrogen predictions
con <- odbcConnectAccess('C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/EC2014.mdb')
   N2002<- sqlQuery(con, "SELECT tblJoinFlowline_WBID_InOut.* FROM tblJoinFlowline_WBID_InOut")

#add to EC.sqlite
  dbWriteTable(conn=EC,name="tblN2002",value=N2002,row.names=FALSE,overwrite=TRUE)
```

**Get tblJoinFlowline_WBID_InOut from MRB1.mdb**
* This table links the NHDplus flowlines COMID to the WBID
* Specifies whether a flowline is an input, output, or internal flowline
* Useful for aggregating the SPARROW nitrogen predictions to lakes.
* add table to EC.sqlite

```{r tblJoinFlowline_WBID_InOut, include=FALSE, echo=FALSE, cache=TRUE}
#read data from EC2014.mdb
  con <- odbcConnectAccess('C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/EC2014.mdb')
   tblJoinFlowline_WBID_InOut<- sqlQuery(con, "SELECT tblJoinFlowline_WBID_InOut.* FROM tblJoinFlowline_WBID_InOut")
#rename WB_ID to WBID
  names(tblJoinFlowline_WBID_InOut)[1]<-'WBID'
#add to EC.sqlite
  dbWriteTable(conn=EC,name="tblJoinFlowline_WBID_InOut",value=tblJoinFlowline_WBID_InOut,row.names=FALSE,overwrite=TRUE)
```

**Merge Tables N2002, tblJoinFlowline_WBID_InOut, and tblJoinWBID_WBRchCd
* Th


```{r mergeTables, include=FALSE, echo=FALSE, cache=TRUE}
#read data from EC2014.mdb
  a<-merge(tblJoinFlowline_WBID_InOut,tblJoinWBID_WBRchCd,by='WBID',all=F)
    nrow(tblJoinFlowline_WBID_InOut) #
    nrow(tblJoinWBID_WBRchCd)        #
    nrow(a)
```






