East Coast Sparrow Model Data
========================================================
<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
  knit('./data/ECdata.rmd')
-->

Create EC.sqlite in the "./data/" subdirectory and add the following tables
-------------------------
* East Coast Sparrow data received from Anne Hoos on September 23 2014 as 4 sas files (forBryan_MRB2_Nitrogen_2002vs2020.zip) and saved to C:\Bryan\EPA\Data\Sparrow\EastCoast\Predictions20140923. The datasets are: 
  * Tables "N2002" and "N2020": The East Coast SPARROW model Nitrogen predictions for each flowline for the CMAQ N deposition predictions for years 2002 (mrb2_predict_tn_2002.sas7bdat) and 2020 (mrb2_predict_tn_2020.sas7bdat). 
  * Tables "R2002" and "R2020": The East Coast SPARROW model Nitrogen predictions for 2002 (mrb2_massbalancereservoirs_2002.sas7bdat) and 2020(mrb2_massbalancereservoirs_2020.sas7bdat) aggregated to lakes (based on reachcodes). 
  * Table "WBIDflowlines": For MRB1 only. Links the flowline COMIDs to the MRB1 WBIDs and shows whether the flowline is an input or output to the lake or an internal reach. Imported from tblJoinFlowline_WBID_InOut (MRB1.mdb)
  * Table "reachcodes" links the MRB1 WaterbodyIDs (WBID) to the NHDplus waterbody Reachcodes (WBRchCd). Created (code below) from: MRB1_NHDWaterbody (WaterbodyDatabase.mdb: the MRB1 NHDplus waterbodies showing the COMIDs and Reachcodes) and tblJoinCOMID_WBID (WaterbodyDatabase.mdb: a table to match waterbody COMIDs to the Lake WBIDs)
    
```{r loadPkgs, include=FALSE, echo=FALSE, cache=FALSE}
  options(stringsAsFactors = FALSE, scipen=9) 


  #function to install (if needed) and load R packages by list
    libs<-c('RODBC','RSQLite','sas7bdat') #list of packages to load
  
    installLoad<-function(pck)#user defined function
    {
      if(!pck%in%installed.packages()){install.packages(pck,repos="http://rweb.quant.ku.edu/cran/")}
      require(pck, character.only = TRUE)
    }
  
    lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

```{r N2002, include=FALSE, echo=FALSE, cache=TRUE}
##release some RAM
  gc() #garbage collection: cleans up temp files
#read the SAS file
  N2002<-read.sas7bdat("C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/mrb2_predict_tn_2002.sas7bdat")
#save as an R dataset
  save(N2002,file='N2002.rda')
#add to EC.sqlite
  EC <- dbConnect(SQLite(), dbname='EC.sqlite')
  dbWriteTable(conn=EC,name="N2002",value=N2002,row.names=FALSE,overwrite=TRUE)
#remove file from workspace to free memory
  rm(N2002) 
```

```{r R2002, include=FALSE, echo=FALSE, cache=TRUE}
##release some RAM
  gc() #garbage collection: cleans up temp files
#read the SAS file
  R2002<-read.sas7bdat("C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/mrb2_massbalancereservoirs_2002.sas7bdat")
#save as an R dataset
  save(R2002,file='R2002.rda')
#add to EC.sqlite
  dbWriteTable(conn=EC,name="R2002",value=R2002,row.names=FALSE,overwrite=TRUE)
#remove file from workspace to free memory
  rm(R2002) 
```

```{r N2020, include=FALSE, echo=FALSE, cache=TRUE}
##release some RAM
  gc() #garbage collection: cleans up temp files
#read the SAS file
  N2020<-read.sas7bdat("C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/mrb2_predict_tn_2020.sas7bdat")
#save as an R dataset
  save(N2020,file='N2020.rda')
#add to EC.sqlite
  dbWriteTable(conn=EC,name="N2020",value=N2020,row.names=FALSE,overwrite=TRUE)
#remove file from workspace to free memory
  rm(N2020) 
```

```{r R2020, include=FALSE, echo=FALSE, cache=TRUE}
##release some RAM
  gc() #garbage collection: cleans up temp files
#read the SAS file
  R2020<-read.sas7bdat("C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/mrb2_massbalancereservoirs_2020.sas7bdat")
#save as an R dataset
  save(R2020,file='R2020.rda')
#add to EC.sqlite
  dbWriteTable(conn=EC,name="R2020",value=R2020,row.names=FALSE,overwrite=TRUE)
#remove file from workspace to free memory
  rm(R2020) 
```

```{r WBIDflowlines, include=FALSE, echo=FALSE, cache=TRUE}
#read data from MRB1.mdb
  con <- odbcConnectAccess("c:/bryan/EPA/Data/WaterbodyDatabase/MRB1.mdb")
  WBIDflowlines<- sqlQuery(con, "SELECT tblJoinFlowline_WBID_InOut.* FROM tblJoinFlowline_WBID_InOut")
#rename WB_ID to WBID
  names(WBIDflowlines)[1]<-'WBID'
#add to EC.sqlite
  dbWriteTable(conn=EC,name="WBIDflowlines",value=WBIDflowlines,row.names=FALSE,overwrite=TRUE)
```

```{r reachcodes, include=FALSE, echo=FALSE, cache=TRUE}
#get tblJoinCOMID_WBID & MRB1_NHDWaterbody
  con <- odbcConnectAccess("c:/bryan/EPA/Data/WaterbodyDatabase/WaterbodyDatabase.mdb")
    tblJoinCOMID_WBID<-sqlQuery(con,"SELECT tblJoinCOMID_WBID.WB_ID, tblJoinCOMID_WBID.COMID
                                     FROM tblJoinCOMID_WBID;
          ")
      nrow(tblJoinCOMID_WBID)  #37192
      length(unique(tblJoinCOMID_WBID$WB_ID))  #36274
    MRB1_NHDWaterbody<-sqlQuery(con,"SELECT MRB1_NHDWaterbody.COMID, MRB1_NHDWaterbody.REACHCODE
                                    FROM MRB1_NHDWaterbody;
          ")
      nrow(MRB1_NHDWaterbody)  #37190
  close(con)

# Temporarily add tables to EC.sqlite
  EC <- dbConnect(SQLite(), dbname='EC.sqlite')
  dbWriteTable(conn=EC,name="tblJoinCOMID_WBID",value=tblJoinCOMID_WBID,row.names=FALSE,overwrite=TRUE)
  dbWriteTable(conn=EC,name="MRB1_NHDWaterbody",value=MRB1_NHDWaterbody,row.names=FALSE,overwrite=TRUE)
        #head(dbReadTable(EC, "tblJoinCOMID_WBID")) 
        #nrow(dbReadTable(EC, "tblJoinCOMID_WBID")) #37192
        #head(dbReadTable(EC, "MRB1_NHDWaterbody")) 
        #nrow(dbReadTable(EC, "MRB1_NHDWaterbody")) #37190
#join tblJoinCOMID_WBID & MRB1_NHDWaterbody and write new table to database
  a<-dbGetQuery(EC, "
    SELECT tblJoinCOMID_WBID.WB_ID,tblJoinCOMID_WBID.COMID,MRB1_NHDWaterbody.REACHCODE
    FROM tblJoinCOMID_WBID INNER JOIN MRB1_NHDWaterbody ON tblJoinCOMID_WBID.COMID = MRB1_NHDWaterbody.COMID
    GROUP BY tblJoinCOMID_WBID.WB_ID, MRB1_NHDWaterbody.REACHCODE;
    ")
        length(a$WB_ID) #36306
        length(unique(a$WB_ID)) #36274
    dbWriteTable(conn=EC,name="temp",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "temp"))
      #nrow(dbReadTable(EC, "temp"))  #36306
  #count the number of gridcodes per WBID
    a<-dbGetQuery(EC, "
      SELECT temp.WB_ID,Count(temp.REACHCODE) AS Count
      FROM temp
      GROUP BY temp.WB_ID
      ORDER BY Count(temp.REACHCODE) DESC
    ")
    dbWriteTable(conn=EC,name="temp1",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "temp1"))
  #add gridcode to temp1 and rename columns
    a<-dbGetQuery(EC, "
      SELECT temp.WB_ID AS WBID,temp.COMID,temp.REACHCODE AS WBRchCd,temp1.Count
      FROM temp INNER JOIN temp1 ON temp.WB_ID=temp1.WB_ID
      ORDER BY temp.WB_ID
    ")
    dbWriteTable(conn=EC,name="reachcodes",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "reachcodes"))
  #delete the working tables  
    dbListTables(EC) # List the tables in the database
    dbRemoveTable(EC, "tblJoinCOMID_WBID")
    dbRemoveTable(EC, "MRB1_NHDWaterbody")
    dbRemoveTable(EC, "temp")
    dbRemoveTable(EC, "temp1")
  #read reachcodes into memory to be used in the text below
      reachcodes<-dbGetQuery(EC, "
        SELECT reachcodes.*
        FROM reachcodes
        ")

#head(reachcodes)
```

Create table "reachcodes": join the WBIDs to the GRIDCODES for MRB1
-------------------------
* Gridcodes for each NHDplus waterbody (COMID) are in the table MRB1_NHDWaterbody in EC2014.mdb
* The waterbody COMIDs and Gridcodes need to be joined to the WBIDs in tblJoinCOMID_WBID
* These tables are imported into a new SQLite databse (EC.sqlite)
* SQL is used to 
  * read the tables from EC.sqlite, join them and write them to a new table.
  * identify the WBIDs with multiple GRIDcodes 
  * create a new table with WBID, COMID, GRIDCODE, and the Count
  * delete the tables temp, temp1, tblJoinCOMID_WBID & MRB1_NHDWaterbody
* we end with a single table (reachcodes) in EC.sqlite
* Evaluate reachcodes
  * `r nrow(reachcodes)` observations
  * `r length(unique(reachcodes$WBID))` unique WBIDs 
  * `r length(unique(reachcodes$WBRchCd))` unique Reachcodes
  * `r nrow(reachcodes[is.na(reachcodes$WBRchCd),])` COMIDs with missing reachcodes; most of these are not really lakes but there is one is a subWaterbody of lake Champlain (WBID=22302965) with a missing reachcode.
  * `r a<-reachcodes[reachcodes$Count>1,'WBID'];length(a)` observations with multiple reachcodes representing `r length(unique(a))` WBID lakes
* Below is a list of the lakes with multiple reachcodes.  These should be removed from the anlaysis.

```{r listLakes, include=TRUE, echo=FALSE, cache=FALSE}
  reachcodes[reachcodes$Count>1,]
```

Data Definitions for reachcodes
-------------------------
**Field** | **Definition**
------------- | -------------
**WBID** | Unique Waterbody ID
**COMID** | NHDplus Unique Waterbody ID-sometimes there are multiple COMID for a single WBID
**WBRchCd** | NHDplus Waterbody Reachcode associated with each COMID; Note this was converted to numeric so leading Zero was lost.  Leading Zero can be added back or used as numeric.
**Count** | Number of WBRchCds associated with the WBID-for analysis keep Count==1 

