East Coast Sparrow Model Compare Methods to Aggregate Loads to Lakes-Data Step
========================================================
Stuff to do 
-------------------------
* ready to go 20140930
* But, R2002 does not have a field for incremental load.  I can still compare my method for calculating inflow and outflow loads but not total lake concentrations.

<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  setwd('data')
  library(knitr)
  knit('ECcompareSQL.rmd')
  
Some useful RSQLite commands
      EC<-dbConnect(SQLite(), dbname='ECcompare.sqlite')
      dbListTables(EC)                         # List the tables in the database
      dbListFields(EC, "reachcodes")               # List the columns in a table
      dbReadTable(EC, "N2002")                # Display the data in a table method1
      dbGetQuery(EC, "SELECT * from flowlines")   # Display the data in a table method2
-->

Create ECcompare.sqlite in the "./data/" subdirectory and add the following tables
-------------------------
* **Table N2002**: Nitrogen predictions for flowlines in MRB1; Incremental loads reported without decay. Original data received from Anne Hoos on September 29 2014 (Sasdataset: "predict_tn_2002_mrb1_exc_incdeca.sas7bdat"" in "Corrected_predictandmassbal_tn_2002_mrb1_excludeincdecay_noatmsec.zip") and saved to C:\Bryan\EPA\Data\Sparrow\EastCoast\Predictions20140923. File read directly into R and saved as a table in ECcompare.sqlite. NOTE:Only a small subset of the available data fields are kept to reduce the size of the database (see ECcompareDataDefinitions.xls).
* **Table R2002**: Nitrogen predictions MRB1 (N2002) aggregated to lakes by reachcode (WBRchCd). Original data received from Anne Hoos on September 29 2014 (Sasdataset: "mrb1_massbalancereservoirs_2002.sas7bdat" in “Corrected_predictandmassbal_tn_2002_mrb1_excludeincdecay_noatmsec.zip”) and saved to C:\Bryan\EPA\Data\Sparrow\EastCoast\Predictions20140923. File read directly into R and saved as a table in ECcompare.sqlite. NOTE:Only a small subset of the available data fields are kept to reduce the size of the database (see ECcompareDataDefinitions.xls). 
* **Table "flowlines**": Links the flowline COMIDs to the MRB1 WBIDs and shows whether the flowline is an input or output to the lake or an internal reach. Imported from tblJoinFlowline_WBID_InOut (MRB1.mdb)
* **Table "reachcodes"** links the MRB1 WaterbodyIDs (WBID) to the NHDplus waterbody Reachcodes (WBRchCd). Created (code below) from: MRB1_NHDWaterbody (WaterbodyDatabase.mdb: the MRB1 NHDplus waterbodies showing the COMIDs and Reachcodes) and tblJoinCOMID_WBID (WaterbodyDatabase.mdb: a table to match waterbody COMIDs to the Lake WBIDs). The reachcodes are used to compare values to R2002.
    
```{r loadPkgs, include=FALSE, echo=FALSE, cache=TRUE}
#options
  options(stringsAsFactors = FALSE, scipen=9) 

  #function to install (if needed) and load R packages by list
    libs<-c('RODBC','RSQLite','sas7bdat') #list of packages to load
  
    installLoad<-function(pck)#user defined function
    {
      if(!pck%in%installed.packages()){install.packages(pck,repos="http://rweb.quant.ku.edu/cran/")}
      require(pck, character.only = TRUE)
    }
  
    lapply(libs,function(x) installLoad(x))  #Load/Install require packages
#data source for sas files
  EC<-dbConnect(SQLite(), dbname='ECcompare.sqlite')
```

```{r N2002, include=FALSE, echo=FALSE, cache=TRUE}
##release some RAM
  gc() #garbage collection: cleans up temp files
#read the SAS file
  N2002<-read.sas7bdat("C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/predict_tn_2002_mrb1_exc_incdeca.sas7bdat")
#select the fields to keep
  N2002<-N2002[,c('comid','WBRchCd','PLOAD_TOTAL','PLOAD_INC_TOTAL','concentration')]
#rename WBRchCd to WBRCHCD
  names(N2002)[2]<-'WBRCHCD'
#add to ECcompare.sqlite
  EC <- dbConnect(SQLite(), dbname='ECcompare.sqlite')
  dbWriteTable(conn=EC,name="N2002",value=N2002,row.names=FALSE,overwrite=TRUE)
#remove file from workspace to free memory
  rm(N2002) 
#view file
  #head(dbReadTable(EC, "N2002"))
```

```{r R2002, include=FALSE, echo=FALSE, cache=TRUE}
##release some RAM
  gc() #garbage collection: cleans up temp files
#read the SAS file
  R2002<-read.sas7bdat("C:/Bryan/EPA/Data/Sparrow/EastCoast/Predictions20140923/mrb1_massbal_2002_exc_incdeca.sas7bdat")
#select the fields to keep
  R2002<-R2002[,c('WBRCHCODE','TN_CONC_inload','TN_massdecay','TOT_CFS',
                  'TN_LOAD_outflow','TN_LOAD_inflow','TN_CONC_inlake')]
#add to ECcompare.sqlite
  dbWriteTable(conn=EC,name="R2002",value=R2002,row.names=FALSE,overwrite=TRUE)
#remove file from workspace to free memory
  rm(R2002) 
```

```{r flowlines, include=FALSE, echo=FALSE, cache=TRUE}
#read data from MRB1.mdb
  con <- odbcConnectAccess("c:/bryan/EPA/Data/WaterbodyDatabase/MRB1.mdb")
  flowlines<- sqlQuery(con, "SELECT tblJoinFlowline_WBID_InOut.* FROM tblJoinFlowline_WBID_InOut")
#rename WB_ID to WBID
  names(flowlines)[1]<-'WBID'
#add to ECcompare.sqlite
  dbWriteTable(conn=EC,name="flowlines",value=flowlines,row.names=FALSE,overwrite=TRUE)
```

```{r reachcodes, include=FALSE, echo=FALSE, cache=TRUE}
#get tblJoinCOMID_WBID & MRB1_NHDWaterbody
  con <- odbcConnectAccess("c:/bryan/EPA/Data/WaterbodyDatabase/WaterbodyDatabase.mdb")
    tblJoinCOMID_WBID<-sqlQuery(con,"SELECT tblJoinCOMID_WBID.WB_ID, tblJoinCOMID_WBID.COMID
                                     FROM tblJoinCOMID_WBID;
          ")
      nrow(tblJoinCOMID_WBID)  #37192
      length(unique(tblJoinCOMID_WBID$WB_ID))  #36274
    MRB1_NHDWaterbody<-sqlQuery(con,"SELECT MRB1_NHDWaterbody.COMID, MRB1_NHDWaterbody.REACHCODE
                                    FROM MRB1_NHDWaterbody;
          ")
      nrow(MRB1_NHDWaterbody)  #37190
  close(con)

# Temporarily add tables to ECcompare.sqlite
  EC <- dbConnect(SQLite(), dbname='ECcompare.sqlite')
  dbWriteTable(conn=EC,name="tblJoinCOMID_WBID",value=tblJoinCOMID_WBID,row.names=FALSE,overwrite=TRUE)
  dbWriteTable(conn=EC,name="MRB1_NHDWaterbody",value=MRB1_NHDWaterbody,row.names=FALSE,overwrite=TRUE)
        #head(dbReadTable(EC, "tblJoinCOMID_WBID")) 
        #nrow(dbReadTable(EC, "tblJoinCOMID_WBID")) #37192
        #head(dbReadTable(EC, "MRB1_NHDWaterbody")) 
        #nrow(dbReadTable(EC, "MRB1_NHDWaterbody")) #37190
#join tblJoinCOMID_WBID & MRB1_NHDWaterbody and write new table to database
  a<-dbGetQuery(EC, "
    SELECT tblJoinCOMID_WBID.WB_ID,tblJoinCOMID_WBID.COMID,MRB1_NHDWaterbody.REACHCODE
    FROM tblJoinCOMID_WBID INNER JOIN MRB1_NHDWaterbody ON tblJoinCOMID_WBID.COMID = MRB1_NHDWaterbody.COMID
    GROUP BY tblJoinCOMID_WBID.WB_ID, MRB1_NHDWaterbody.REACHCODE;
    ")
        length(a$WB_ID) #36306
        length(unique(a$WB_ID)) #36274
    dbWriteTable(conn=EC,name="temp",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "temp"))
      #nrow(dbReadTable(EC, "temp"))  #36306
  #count the number of gridcodes per WBID
    a<-dbGetQuery(EC, "
      SELECT temp.WB_ID,Count(temp.REACHCODE) AS Count
      FROM temp
      GROUP BY temp.WB_ID
      ORDER BY Count(temp.REACHCODE) DESC
    ")
    dbWriteTable(conn=EC,name="temp1",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "temp1"))
  #add gridcode to temp1 and rename columns
    a<-dbGetQuery(EC, "
      SELECT temp.WB_ID AS WBID,temp.COMID,temp.REACHCODE AS WBRCHCD,temp1.Count
      FROM temp INNER JOIN temp1 ON temp.WB_ID=temp1.WB_ID
      ORDER BY temp.WB_ID
    ")
    dbWriteTable(conn=EC,name="reachcodes",value=a,row.names=FALSE,overwrite=TRUE)
      #head(dbReadTable(EC, "reachcodes"))
  #delete the working tables  
    dbListTables(EC) # List the tables in the database
    dbRemoveTable(EC, "tblJoinCOMID_WBID")
    dbRemoveTable(EC, "MRB1_NHDWaterbody")
    dbRemoveTable(EC, "temp")
    dbRemoveTable(EC, "temp1")
  #read reachcodes into memory to be used in the text below
      reachcodes<-dbGetQuery(EC, "
        SELECT reachcodes.*
        FROM reachcodes
        ")

#head(reachcodes)
```

Create table "reachcodes": join the WBIDs to the GRIDCODES for MRB1
-------------------------
* Gridcodes for each NHDplus waterbody (COMID) are in the table MRB1_NHDWaterbody in EC2014.mdb
* The waterbody COMIDs and Gridcodes need to be joined to the WBIDs in tblJoinCOMID_WBID
* These tables are imported into a new SQLite databse (ECcompare.sqlite)
* SQL is used to 
  * read the tables from ECcompare.sqlite, join them and write them to a new table.
  * identify the WBIDs with multiple GRIDcodes 
  * create a new table with WBID, COMID, GRIDCODE, and the Count
  * delete the tables temp, temp1, tblJoinCOMID_WBID & MRB1_NHDWaterbody
* we end with a single table (reachcodes) in ECcompare.sqlite
* Evaluate reachcodes
  * `r nrow(reachcodes)` observations
  * `r length(unique(reachcodes$WBID))` unique WBIDs 
  * `r length(unique(reachcodes$WBRCHCD))` unique Reachcodes
  * `r nrow(reachcodes[is.na(reachcodes$WBRCHCD),])` COMIDs with missing reachcodes; most of these are not really lakes but there is one (a subWaterbody of lake Champlain; WBID=22302965) with a missing reachcode.
  * `r a<-reachcodes[reachcodes$Count>1,'WBID'];length(a)` observations with multiple reachcodes representing `r length(unique(a))` WBID lakes
* Below is a list of the lakes with multiple reachcodes (reachcodes$Count>1).  These should be removed from the anlaysis.

```{r listLakes, include=TRUE, echo=FALSE, cache=TRUE}
  reachcodes[reachcodes$Count>1,]
```

Data Definitions for table "N2002"
-------------------------
**Field** | **Definition**
------------- | -------------
**comid** | Stream Reach Identifier from NHDplus
**WBRCHCD** | Unique identifier for a reservoir or lake,  from nhdwaterbody dbf file Aug 2011
**PLOAD_TOTAL** | Total Flux (kg/yr)
**PLOAD_INC_TOTAL** | Total Incremental (Inc) Flux (kg/yr)
**concentration** | Mean Flow-Weighted Concentration (mg/L)


Data Definitions for table "R2002"
-------------------------
**Field** | **Definition**
------------- | -------------
**TN_CONC_inload** | Tributary inflow Nitrogen Concentration, mg/L  -  this is value reported in USGS DS 820 as "Concentration of tributary inflow to receiving water", calculated as  (TN_LOAD_inflow) / (TOT_CFS * 28.32 * 1/1000000 * 86400 * 365 ).   Would have preferred to use sum of tributary inflow MAFLOWU as the demonimator but this was difficult to compute;  thus it's a slightly mixed calculation
**TN_massdecay** | Nitrogen mass removed in the lake,  kg/yr, summed from the amount removed in each incremental reservoir segment - this is value reported in USGS DS 820 as "Load assimilated in receiving waterbody"
**TOT_CFS** | Flow (ft3/sec): to convert to m3/yr multiply by 893593
**TN_LOAD_outflow** | Total nitrogen load at downstream node of exit reach, in kg/yr - this is the value reported in USGS DS 820 source share table as "Total load".    Wanted to use TN_LOAD_inflow here, to be consistent with use of trib inflow info in the table that follows, but see explanation at top of this file
**TN_LOAD_inflow** | Total nitrogen load entering lake from all tributaries and adjacent areas,  calculated from mass balance (load at downstream node of exit reach plus load removed in lake)
**TN_CONC_inlake** | Estimated In-Lake Nitrogen Concentration of reservoir  (mg/L), calculated as TN_arealloading_outload / (hload * (1 - RetentCoeff_Nurnberg))


Data Definitions for table "flowlines"
-------------------------
**Field** | **Definition**
------------- | -------------
**WBID** | Unique lake identifier from WaterbodyDatabase.mdb
**SelectLake** | 1=lake not flagged; 0=at least one reach in lake has a problem (missing data, diverision, Nin not equal Nout)
**Reach** | NHDplus identifier (COMID)
**GridCode** | Stream Reach GridCode
**LakeOutflow** | 1 = reach is an outflow; 0 = not an outflow
**ReachInLake** | 1 = reach is internal to lake; 0 = not in lake
**ReachUpstream** | 1 = reach immediately downstream of lake; 0 = not downstream
**ReachDownstream** | 1 = reach immediately upstream of lake; 0 = not upstream
**FlagReach** | 1 = problem with reach 
**LakeFcode** | NHDplus Fcode for Waterbody
**ReachFcode** | NHDplus Fcode for Stream Reach
**HasNPData** | The original MRB1 SPARROW model had N/P data for this reach
**Diversion** | Known diversion in the reach
**NinEqNout** | Nitrogen Inputs to reach = outputs

Data Definitions for table "reachcodes"
-------------------------
**Field** | **Definition**
------------- | -------------
**WBID** | Unique Waterbody ID
**COMID** | NHDplus Unique Waterbody ID-sometimes there are multiple COMID for a single WBID
**WBRCHCD** | NHDplus Waterbody Reachcode associated with each COMID; Note this was converted to numeric so leading Zero was lost.  Leading Zero can be added back or used as numeric.
**Count** | Number of WBRchCds associated with the WBID-for analysis keep Count==1 



