Create Spatial Objects for Waterbodies in HUC regions 01, 02, 03, and 06
========================================================
Stuff to do 
-------------------------
* ready 20141009

<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
First make sure you are in the data directory: 
  setwd('data')
  library(knitr)
  knit('lakesSpatialData.rmd')
  
-->
Spatial Data
-------------------------
* read the nhdwaterbody.shp files for the four regions
* identify and remove duplicate lakes
* merge lakes from the regions
* dissolve based on reachcode-output= SpatialPolygonsDataframe "lakes" with geographic coordinates
* get the centroids and convert to spdf with reachcode (WBRCHCODE) as an attribute. output= SpatialPointsDataframe "lakes" with geographic coordinates
* Three spatial and one tabular datasets are created for the lakes in MRB1 and MRB2; both in geographic coordinates (NAD83)
  * lakes: a spatial Polygons Dataframe with the lake locations based on reachcode (WBRCHCODE)
  * lakesPt: a spatial Points Dataframe with the lake centroids based on reachcode (WBRCHCODE)
  * ECstates: state boundaries (derived from a shapefile from Jane Copeland) for states in MRB1 and MRB2
  * lakeStates: a table matching the 88662 lakes with reachcodes to states (note: lakes in more than 1 state arbitrarily assigned to 1 of the states.)
* data saved in "lakesSpatialData.rda"
* Link to this document: https://github.com/willbmisled/ECsparrow/blob/master/data/lakesSpatialData.md
  
```{r loadPkgs, include=FALSE, echo=FALSE, cache=TRUE}
  options(stringsAsFactors = FALSE, scipen=9) 


  #function to install (if needed) and load R packages by list
    libs<-c('RSQLite','rgdal','maptools','RODBC','maps','rgeos') #list of packages to load
  
    installLoad<-function(pck)#user defined function
    {
      if(!pck%in%installed.packages()){install.packages(pck,repos="http://rweb.quant.ku.edu/cran/")}
      require(pck, character.only = TRUE)
    }
  
    lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

```{r getShapes, include=FALSE, echo=FALSE, cache=TRUE}
#getwd()  #need to be in ./data/
#Get the NHDplus shapefiles for HUC regions 01,02, and 03
  HUC01<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus01/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))
  HUC02<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus02/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))
  HUC03<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus03/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))
  HUC06<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus06/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))

#eliminate lakes where REACHCODE==NA
  table(!is.na(HUC01$REACHCODE))          #F=4250 T=12832
    HUC01<-HUC01[!is.na(HUC01$REACHCODE),]
      nrow(HUC01)                         #12832  
  table(!is.na(HUC02$REACHCODE))          #F=4621 T=15484
    HUC02<-HUC02[!is.na(HUC02$REACHCODE),]
      nrow(HUC02)                         #15484  
  table(!is.na(HUC03$REACHCODE))          #F=23361 T=59829 
    HUC03<-HUC03[!is.na(HUC03$REACHCODE),]
      nrow(HUC03)                         #59829  
  table(!is.na(HUC06$REACHCODE))          #F=478  T=1529 
    HUC06<-HUC06[!is.na(HUC06$REACHCODE),]
      nrow(HUC06)                         #1529  
```

```{r mergeShapes, include=FALSE, echo=FALSE, cache=TRUE}
#check to see if lakes are repeated
  w<-row.names(as(HUC01, "data.frame"))
  x<-row.names(as(HUC02, "data.frame"))
  y<-row.names(as(HUC03, "data.frame"))
  z<-row.names(as(HUC06, "data.frame"))
  
  table(w%in%x)  #no matches
  table(w%in%y)  #no matches
  table(w%in%z)  #no matches
  table(x%in%y)  #no matches
  table(x%in%z)  #no matches
  table(y%in%z)  #no matches

#merge the four layers
  WB<-spRbind(HUC01,HUC02)
  WB<-spRbind(WB,HUC03)
  WB<-spRbind(WB,HUC06)
    nrow(WB)==nrow(HUC01)+nrow(HUC02)+nrow(HUC03)+nrow(HUC06)  #89674
```

```{r dissolveShapes, include=FALSE, echo=FALSE, cache=TRUE}
#dissolve the polygons by REACHCODE
  length(WB$REACHCODE) #89674
  length(unique(WB$REACHCODE)) #88662
    lakes<-unionSpatialPolygons(WB, WB$REACHCODE)
      nrow(lakes) #88662  
#convert to SpatialPolygonsDataframe with reachcode as an attribute
  data<-data.frame(WBRCHCODE=row.names(lakes))  #the reachcodes
    row.names(data)<-row.names(lakes)
  lakes<-SpatialPolygonsDataFrame(lakes,data=data)
    nrow(lakes) #88662 
```

```{r fixHoles, include=FALSE, echo=FALSE, cache=TRUE}
#Associate lake holes (islands) with the correct polygon. 
  slot(lakes, "polygons") <- lapply(slot(lakes, "polygons"), checkPolygonsHoles)
```

```{r points, include=FALSE, echo=FALSE, cache=TRUE}
#create spatial points dataframe of lake centroids
  lakesPt<-gCentroid(lakes, byid=TRUE, id = NULL)  #reachcodes are the row.names for the spatialpoints object
#convert to spatial points data.frame
  coords<-coordinates(lakesPt)  #the centroids
  data<-data.frame(WBRCHCODE=row.names(lakesPt))  #the reachcodes
      row.names(data)<-row.names(lakesPt)
  lakesPt<-SpatialPointsDataFrame(coords, data)  #convert to spdf
    proj4string(lakesPt)<-CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06") #set crs
  

#Check to see if all of the reachcodes for lakes with SPARROW data are included.
  #Get the SPARROW lakes reachcodes.
    EC<-dbConnect(SQLite(), dbname='EClakes.sqlite')
      a<-dbGetQuery(EC, "SELECT WBRCHCODE from R2002m1") 
      b<-dbGetQuery(EC, "SELECT WBRCHCODE from R2002m2")
    RC<-rbind(a,b)
  #compare reachcodes to lakesPt
    test<-RC$WBRCHCODE%in%lakesPt$WBRCHCODE
    table(test) #FALSE=43 TRUE=41523
    a<-RC[!test,]
```

Missing reachcodes 
-------------------------
* We have spatial data (lakesPt or lakes) for all but 43 of the lakes in EC aggregated lakes files (EClakes.sqlite)
* The documentation of this database shows that some reachcodes were derived for lakes in MRB2.  
* These are probably the missing lakes.
* Here is a list of the missing reachcodes.

```{r missing, include=TRUE, echo=FALSE, cache=TRUE}
RC[!test,]
```

```{r statesMap, include=FALSE, echo=FALSE, cache=TRUE}
#get us states (Data source = Jane Copeland)
  US50wgs84<- readShapePoly('L:/Public/Milstead_Lakes/GIS/StateBoundaries/States.shp', 
                       proj4string=CRS("+proj=longlat +datum=WGS84"))   
#reproject to NAD83
  US50nad83<-spTransform(US50wgs84,CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))
#crop file to EC states
  keep<-which(US50wgs84@data$STATE_ABBR%in%c('AL','CT','DE','DC','FL','GA','KY','LA','ME','MD','MA','MS',
                                              'NH','NJ','NY','NC','PA','RI','SC','TN','VT','VA','WV'))
  ECstates<-US50nad83[keep,]
    #plot(lakesPt,pch=16,cex=.001,col='red')
    #plot(ECstatesNAD83a,add=TRUE)
#delete files to free memory
  rm(US50wgs84)
  rm(US50nad83)
#save the data
  save(ECstates,lakes,lakesPt,file='lakesSpatialData.rda')
```

```{r statesTable, include=FALSE, echo=FALSE, cache=TRUE}  
#assign state names to table TSN
  #Free memory-this is a memory hawg.
    rm(list=ls()) #clear workspace
  #load the data
    load(file='lakesSpatialData.rda')
  #use loop to do the overlay
    # this can't be done all at once so split into bins and then rbind the results
      #' x is the number of lakes to do in each bin
      #' lakes is the lakes spatial object
      #' states is the states spatial object
    fOver<-function(x=1000,Lakes=lakesPt,States=ECstates) {  
      k<-floor(nrow(Lakes)/x) #number of loops
      a<-as.data.frame(over(Lakes[1:x,],States)) #start database
      a[(x+1):nrow(Lakes),]<-NA   #add blank rows
      row.names(a)<-row.names(Lakes) #these are the lake reachcodes; need to keep them
      #loop to add more lakes       
        for(i in 2:k){
          start<-((i-1)*x)+1
          end<-start+x-1
          a[start:end,]<-over(Lakes[start:end,],States) #add lakes
        }
      #add last lakes
        start<-(k*x)+1
        end<-nrow(Lakes)
        a[start:end,]<-over(Lakes[start:end,],States) #add lakes
      #output results
        return(a)
    } 
  
  #get state/lake matches
    #overlay the centroids first
      lakePtStates<-fOver(x=1000,Lakes=lakesPt,States=ECstates) 
        table(is.na(lakePtStates$STATE_ABBR)) #388 of 88662 not assigned
      #write nonNA data to lakeStates
        lakeStates<-na.exclude(data.frame(WBRCHCODE=row.names(lakePtStates),
                    State=lakePtStates$STATE_NAME,ST=lakePtStates$STATE_ABBR))        
    #for lake centroids not assigned overlay the lake polgyons
      all.equal(row.names(lakes),row.names(lakesPt))  #TRUE so lakes are in the same order
      keep<-is.na(lakePtStates$STATE_ABBR) #missing lakes
      lakes1<-lakes[keep,] #polygons of missing lakes
        lakePtStates1<-as.data.frame(over(lakes1,ECstates)) 
        table(is.na(lakePtStates1)) #11 of 388 not assigned
      #add nonNA data to lakeStates
        a<-na.exclude(lakePtStates1) #exclude non-assigned lakes
        names(a)<-'Index' #results are the row indices of ECstates 
        a$WBRCHCODE<-row.names(a)
        b<-ECstates@data #copy ECstates data
        b$Index<-1:nrow(b) #add merge field "Index"
        q<-merge(a,b,by='Index',all.x=T)[,c("WBRCHCODE","STATE_NAME","STATE_ABBR")] #merge data
          names(q)<-c("WBRCHCODE","State","ST")
        table(is.na(q$ST)) #false=377
        lakeStates<-rbind(lakeStates,q)
    #list of lakes to check visually
        chkLakes<-lakePtStates1 #copy before modifying
        chkLakes$WBRCHCODE<-row.names(chkLakes)
        chkLakes<-chkLakes[is.na(chkLakes),'WBRCHCODE']
    #add chkLakes to lakeStates
      a<-unique(lakeStates[-1])
      b<-data.frame(WBRCHCODE=chkLakes,ST=c("MA","MA","MD","VA","VA","VA","VA","NC","NC","NC","FL"))
      q<-merge(b,a,by='ST',all.x=TRUE)
      chkLakes<-q[,c(2,3,1)]
      lakeStates<-rbind(lakeStates,chkLakes)
        nrow(lakeStates)
    #save the data
      save(ECstates,lakes,lakesPt,file='lakesSpatialData.rda') 
```

Assign Lakes to States
-------------------------
* use the "over" function (packages 'sp' and 'rgeos') to overlay lake Centroids on states
* this returns a df with 88274 of the 88662 lakes assigned to states
* NOTE: if a lake is more than 1 state it is arbitrarily assigned to one of the states (probably the one with the lowest index value)
* The 388 lakes that were not assigned are probably on the coast side of the state and just outside of the boundary.
* Next step is to overlay this polygons of the missing lakes on the states.
* This returns a df with 377 of the 388 missing lakes assigned to a state
* This leaves 11 lakes unassigned-for completeness I will check these visually and assign them to states.
* These were viewed individually in ArcGIS and assigned to states.
* Visually verified lakes include:

```{r chkLakes, include=TRUE, echo=FALSE, cache=TRUE} 
chkLakes
```

Data Definitions
-------------------------
**SpatialPolygonsDataFrame "Lakes" and SpatialPointsDataFrame "LakesPt"**
* 88662 observations
* Geographic Coordinates datum=NAD83 (proj4string="+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06")

**Field** | **Definition**
------------- | -------------
**WBRCHCODE** | Waterbody Reachcode, unique identifier for a reservoir or lake from the attribute 'REACHCODE' in the shapefile(s) NHDwaterbody.  

**DataFrame "lakeStates"** 
* 88662 observations
* Geographic Coordinates datum=NAD83 (proj4string="+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06")

**Field** | **Definition**
------------- | -------------
**WBRCHCODE** | Waterbody Reachcode, unique identifier for a reservoir or lake from the attribute 'REACHCODE' in the shapefile(s) NHDwaterbody. 
**State** | State names long form
**ST** | Two letter state abbreviation.

**SpatialPolygonsDataFrame "ECstates"** 
* 23 observations
* Geographic Coordinates datum=NAD83 (proj4string="+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06")
* Attributes appear to be summary stats from the 2010 census.  Fields include:

```{r ECstates, include=FALSE, echo=FALSE, cache=TRUE} 
names(ECstates)
```


