Create Spatial Objects for Waterbodies in HUC regions 01, 02, 03, and 06
========================================================
Stuff to do 
-------------------------
* ready 20140930

<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
First make sure you are in the data directory: 
  setwd('data')
  library(knitr)
  knit('lakesSpatialData.rmd')
  
-->


Spatial Data
-------------------------
* read the nhdwaterbody.shp files for the four regions
* identify and remove duplicate lakes
* merge lakes from the regions
* dissolve based on reachcode-output= SpatialPolygonsDataframe "lakes" with geographic coordinates
* get the centroids and convert to spdf with reachcode (WBRCHCODE) as an attribute. output= SpatialPointsDataframe "lakes" with geographic coordinates
* save data in "lakesSpatialData.rda"
* two spatial datasets are created for the lakes in MRB1 and MRB2; both in geographic coordinates
  * lakes: a spatial Polygons Dataframe with the lake locations based on reachcode (WBRCHCODE)
  * lakesPt: a spatial Points Dataframe with the lake centroids based on reachcode (WBRCHCODE)
* Link to this document: https://github.com/willbmisled/ECsparrow/blob/master/data/lakesSpatialData.md
  
```{r loadPkgs, include=FALSE, echo=FALSE, cache=TRUE}
  options(stringsAsFactors = FALSE, scipen=9) 


  #function to install (if needed) and load R packages by list
    libs<-c('RSQLite','rgdal','maptools','RODBC') #list of packages to load
  
    installLoad<-function(pck)#user defined function
    {
      if(!pck%in%installed.packages()){install.packages(pck,repos="http://rweb.quant.ku.edu/cran/")}
      require(pck, character.only = TRUE)
    }
  
    lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

```{r getShapes, include=FALSE, echo=FALSE, cache=TRUE}
#getwd()  #need to be in ./data/
#Get the NHDplus shapefiles for HUC regions 01,02, and 03
  HUC01<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus01/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))
  HUC02<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus02/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))
  HUC03<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus03/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))
  HUC06<-readShapePoly('C:/Bryan/EPA/Data/GIS/NHDPlus/NHDPlus06/hydrography/nhdwaterbody.shp',IDvar="COMID", 
                      proj4string=CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,06"))

#eliminate lakes where REACHCODE==NA
  table(!is.na(HUC01$REACHCODE))          #F=4250 T=12832
    HUC01<-HUC01[!is.na(HUC01$REACHCODE),]
      nrow(HUC01)                         #12832  
  table(!is.na(HUC02$REACHCODE))          #F=4621 T=15484
    HUC02<-HUC02[!is.na(HUC02$REACHCODE),]
      nrow(HUC02)                         #15484  
  table(!is.na(HUC03$REACHCODE))          #F=23361 T=59829 
    HUC03<-HUC03[!is.na(HUC03$REACHCODE),]
      nrow(HUC03)                         #59829  
  table(!is.na(HUC06$REACHCODE))          #F=478  T=1529 
    HUC06<-HUC06[!is.na(HUC06$REACHCODE),]
      nrow(HUC06)                         #1529  
```

```{r mergeShapes, include=FALSE, echo=FALSE, cache=TRUE}
#check to see if lakes are repeated
  w<-row.names(as(HUC01, "data.frame"))
  x<-row.names(as(HUC02, "data.frame"))
  y<-row.names(as(HUC03, "data.frame"))
  z<-row.names(as(HUC06, "data.frame"))
  
  table(w%in%x)  #no matches
  table(w%in%y)  #no matches
  table(w%in%z)  #no matches
  table(x%in%y)  #no matches
  table(x%in%z)  #no matches
  table(y%in%z)  #no matches

#merge the four layers
  WB<-spRbind(HUC01,HUC02)
  WB<-spRbind(WB,HUC03)
  WB<-spRbind(WB,HUC06)
    nrow(WB)==nrow(HUC01)+nrow(HUC02)+nrow(HUC03)+nrow(HUC06)  #89674
```

```{r dissolveShapes, include=FALSE, echo=FALSE, cache=TRUE}
#dissolve the polygons by REACHCODE
  length(WB$REACHCODE) #89674
  length(unique(WB$REACHCODE)) #88662
    lakes<-unionSpatialPolygons(WB, WB$REACHCODE)
      nrow(lakes) #88662  
#convert to SpatialPolygonsDataframe with reachcode as an attribute
  data<-data.frame(WBRCHCODE=row.names(lakes))  #the reachcodes
    row.names(data)<-row.names(lakes)
  lakes<-SpatialPolygonsDataFrame(lakes,data=data)
    nrow(lakes) #88662 
```

```{r fixHoles, include=FALSE, echo=FALSE, cache=TRUE}
#Associate lake holes (islands) with the correct polygon. 
  slot(lakes, "polygons") <- lapply(slot(lakes, "polygons"), checkPolygonsHoles)
```

```{r points, include=FALSE, echo=FALSE, cache=TRUE}
#create spatial points dataframe of lake centroids
  lakesPt<-gCentroid(lakes, byid=TRUE, id = NULL)  #reachcodes are the row.names for the spatialpoints object
#convert to spatial points data.frame
  coords<-coordinates(lakesPt)  #the centroids
  data<-data.frame(WBRCHCODE=row.names(lakesPt))  #the reachcodes
      row.names(data)<-row.names(lakesPt)
  lakesPt<-SpatialPointsDataFrame(coords, data)  #convert to spdf
#Check to see if all of the reachcodes for lakes with SPARROW data are included.
  #Get the SPARROW lakes reachcodes.
    EC<-dbConnect(SQLite(), dbname='EClakes.sqlite')
      a<-dbGetQuery(EC, "SELECT WBRCHCODE from R2002m1") 
      b<-dbGetQuery(EC, "SELECT WBRCHCODE from R2002m2")
    RC<-rbind(a,b)
  #compare reachcodes to lakesPt
    test<-RC$WBRCHCODE%in%lakesPt$WBRCHCODE
    table(test) #FALSE=43 TRUE=41523
    a<-RC[!test,]
    

#save the data
  save(lakes,lakesPt,file='lakesSpatialData.rda')
```

Missing reachcodes 
-------------------------
* We have spatial data (lakesPt or lakes) for all but 43 of the lakes in EC aggregated lakes files (EClakes.sqlite)
* The documentation of this database shows that some reachcodes were derived for lakes in MRB2.  
* These are probably the missing lakes.
* Here is a list of the missing reachcodes.

```{r missing, include=TRUE, echo=FALSE, cache=TRUE}
RC[!test,]
```
