East Coast SPARROW: Lake Trophic State Change Analysis for the CMAQ Scenario
========================================================
Stuff to do 
-------------------------
* ready to go 20141009


<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
First make sure you are in the analysis directory: 
  getwd()
  setwd('..')
  setwd('analysis')
  library(knitr)
  knit('trophicStateCMAQ.rmd')
  
Some useful RSQLite commands
      EC<-dbConnect(SQLite(), dbname='../data/EClakes.sqlite')
      dbListTables(EC)                         # List the tables in the database
      dbListFields(EC, "R2002m2")               # List the columns in a table
      dbReadTable(EC, "R2002m2")                # Display the data in a table method1
      dbGetQuery(EC, "SELECT * from R2020m1")   # Display the data in a table method2
-->

Introduction
-------------------------
* The East Coast SPARROW model predicts nitrogen loads to streams in MRB1 and MRB2 (HUC regions 01, 02, 03, and 06)
* Information on the East Coast Model can be found here:
  * Moorman, M.C., Hoos, A.B., Bricker, S.B., Moore, R.B., García, A.M., and Ator, S.W., 2014, Nutrient load summaries for major lakes and estuaries of the Eastern United States, 2002: U.S. Geological Survey Data Series 820, 94 p.,  http://dx.doi.org/10.3133/ds820.
* Estimates of atmospheric deposition of nitrogen are derived from the CMAQ (Community Multi-scale Air Quality Model) model (http://www.epa.gov/AMD/Research/RIA/cmaq.html)
* The CMAQ model has:
  * Interpolated N deposition based on 2002 data
  * Predicted N deposition for 2020 following changes to air rules
* Our goal is to evaluate changes in lake trophic state between 2002 and 2020 based on the EC SPARROW model prediction for nitrogen loads to the lakes in MRB1 and MRB2
* Link to this document: https://github.com/willbmisled/ECsparrow/blob/master/analysis/trophicStateCMAQ.md

Objectives
-------------------------
* Estimate trophic state of lakes based on 2002 conditions
* Estimate trophic state of lakes based on 2020 conditions
* Analyze changes in trophic state between 2002 and 2020
* Map changes in trophic state

Data Sources
-------------------------
* The 2002 and 2020 sparrow N predictions aggregated to lakes for MRB1 and MRB2 are in: '/data/EClakes.sqlite'. 
* Data Definitions: https://github.com/willbmisled/ECsparrow/blob/master/data/EClakesSQL.md
* Includes the following tables:
  * **R2002m1**: N predictions based on 2002 conditions for lakes in MRB1
  * **R2020m1**: N predictions based on 2020 conditions for lakes in MRB1
  * **R2002m2**: N predictions based on 2002 conditions for lakes in MRB2
  * **R2020m2**: N predictions based on 2020 conditions for lakes in MRB2
* Polygon (lakes) and point (lakesPt) locations of the lakes are available in '/data/lakesSpatialData.rda'.  
* ECstates boundaries are available in '/data/ECstates.rda'.  
* Data Definitions for spatial data: https://github.com/willbmisled/ECsparrow/blob/master/data/lakesSpatialData.md

```{r loadPkgs, include=FALSE, echo=FALSE, cache=FALSE}
  options(stringsAsFactors = FALSE, scipen=9) 

  #function to install (if needed) and load R packages by list
    libs<-c('RSQLite','rgdal','maptools','RODBC') #list of packages to load
  
    installLoad<-function(pck)#user defined function
    {
      if(!pck%in%installed.packages()){install.packages(pck,repos="http://rweb.quant.ku.edu/cran/")}
      require(pck, character.only = TRUE)
    }
  
    lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

```{r data, include=FALSE, echo=FALSE, cache=FALSE}
#get the lakes spatial data
    load('../data/lakesSpatialData.rda')
    load('../data/ECstates.rda')

#transform ECstates and lakesPT to albers 
  #ESRI USA Contiguous Albers Equal Area Conic (used by MRB1 WBIDLakes as AlbersX and AlbersY)
    Albers<-CRS('+proj=aea +x_0=0 +y_0=0 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +units=m +datum=NAD83')
    lakesPtAlbers<-spTransform(lakesPt,Albers)
    statesAlbers<-spTransform(ECstates,Albers)


#get the estimates of N loads under the CMAQ Scenarios
  #connect to EClakes.sqlite
      EC<-dbConnect(SQLite(), dbname='../data/EClakes.sqlite')
      dbListTables(EC) # List the tables in the database
      dbListFields(EC,"R2002m1")               # List the columns in a table
      dbListFields(EC,"R2002m2")               # List the columns in a table
  #get the 2002 estimates
    a<-dbGetQuery(EC, "SELECT WBRCHCODE,TOT_CFS,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2002m1") 
    b<-dbGetQuery(EC, "SELECT WBRCHCODE,TOT_CFS,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2002m2") 
    N2002<-rbind(a,b)
  #get the 2020 estimates
    a<-dbGetQuery(EC, "SELECT WBRCHCODE,TOT_CFS,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2020m1") 
    b<-dbGetQuery(EC, "SELECT WBRCHCODE,TOT_CFS,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2020m2") 
    N2020<-rbind(a,b)

#add outflow concentrations following same methods as Anne and Rich (see EClakesSQL.Rmd)
  N2002$TN_CONC_outload<-N2002$TN_LOAD_outflow/ (N2002$TOT_CFS * 28.32 * 1/1000000 * 86400 * 365 )
  N2020$TN_CONC_outload<-N2020$TN_LOAD_outflow/ (N2020$TOT_CFS * 28.32 * 1/1000000 * 86400 * 365 )
  
#Check for missing values
  table(is.na(N2002$TN_CONC_outload))   #no missing values
  table(is.na(N2020$TN_CONC_outload))   #no missing values
  table(is.na(N2002$TN_CONC_inload))    #no missing values
  table(is.na(N2020$TN_CONC_inload))    #no missing values
  table(is.na(N2002$TN_CONC_inlake))    #no missing values
  table(is.na(N2020$TN_CONC_inlake))    #no missing values
  
#Check number of observations and order of reachcode
  nrow(N2002)  #41566
  nrow(N2020)  #41566
  all.equal(N2002$WBRCHCODE,N2020$WBRCHCODE) #same order
```

```{r TSNtable, include=FALSE, echo=FALSE, cache=FALSE}
#select TS criteria
  criteriaNLA<-c(.35,.75,1.4)  #mg/l
#function to assign Trophic State in lakes based on Nitrogen thresholds
  #' criteria=thresholds between the trophic states; default=criteriaNLA
  #' conc=which of 3 estimates of lake concentration to use; default=N2002$TN_CONC_inlake
  estTSN<-function(criteria=criteriaNLA,conc=L2002$TN_CONC_outload){ 
    TSN<-factor(rep('Oligo',length(conc)),ordered=T,levels=c("Oligo","Meso","Eu","Hyper"))
    TSN[conc>criteria[1]]<-'Meso'
    TSN[conc>criteria[2]]<-'Eu'
    TSN[conc>criteria[3]]<-'Hyper'
    TSN[is.na(conc)]<-NA
  return(TSN)
  }

#Estimate trophic state for each lake  
  TSN<-data.frame(WBRCHCODE=N2002$WBRCHCODE)
  TSN$out2002<-estTSN(conc=N2002$TN_CONC_outload)
  TSN$out2020<-estTSN(conc=N2020$TN_CONC_outload)
    TSN$outChg<-as.numeric(TSN$out2020)-as.numeric(TSN$out2002)
  TSN$in2002<-estTSN(conc=N2002$TN_CONC_inload)
  TSN$in2020<-estTSN(conc=N2020$TN_CONC_inload)
    TSN$inChg<-as.numeric(TSN$in2020)-as.numeric(TSN$in2002)
  TSN$lake2002<-estTSN(conc=N2002$TN_CONC_inlake)
  TSN$lake2020<-estTSN(conc=N2020$TN_CONC_inlake)
    TSN$lakeChg<-as.numeric(TSN$lake2020)-as.numeric(TSN$lake2002)
   
#create a table of trophic state totals for each TN estimate
  tblTSN<-function(df=TSN,x='out'){
    z<-data.frame(Year=c(2002,2020),Estimate=c(x,x))
    a<-as.data.frame(table(df[,paste(x,'2002',sep='')],useNA='ifany'))[,2]
    b<-as.data.frame(table(df[,paste(x,'2020',sep='')],useNA='ifany'))[,2]
    a<-data.frame(rbind(a,b))
    #add column for missing values 
      if(ncol(a)==4) a[,5]<-0 
    #rename columns
    names(a)<-c('Oligo','Meso','Eu','Hyper','Missing')
    #add info
      z<-cbind(z,a)
    #add column for improved lakes
      chg<-na.exclude(df[,paste(x,'Chg',sep='')])
      z[2,'Improved']<-length(chg[chg<0])
    #add column for deteriorated lakes
      chg<-na.exclude(df[,paste(x,'Chg',sep='')])
      z[2,'Worse']<-length(chg[chg>0])    
    return(z)
  } 

#build the table
  a<-rbind(tblTSN(x='out'),
  tblTSN(x='in'),
  tblTSN(x='lake'))
  row.names(a)<-1:6
  TSNtable<-a
  #TSNtable
```

```{r TSNstates, include=FALSE, echo=FALSE, cache=FALSE}
#add states to TSN
  #get the MRB1 lake state matches from Anne
    #connect to EClakes.sqlite
        EC<-dbConnect(SQLite(), dbname='../data/EClakes.sqlite')
    #get the state fips codes for MRB1 supplied by Anne
      Fips<-dbGetQuery(EC, "SELECT WBRCHCODE,FIPS_ST from R2002m2") 
      Codes<-data.frame(FIPS_ST=unique(Fips$FIPS_ST),ST_Anne=c("VA","NC",NA,"SC","GA","FL","AL","MS","TN","LA","KY"))
      Fips<-merge(Fips,Codes,by='FIPS_ST',all.x=TRUE)
      table(is.na(Fips$ST_Anne)) #100 missing values
      Fips<-na.exclude(Fips) #remove missing values
    #merge Fips and lakeStates to compare values for ST
      Compare<-merge(Fips,lakeStates,by='WBRCHCODE',all.x=TRUE)
    #compare values for ST
      Match<-Compare$ST==Compare$ST_Anne
      table(Match)   #F=34 T=23510
#merge TSN and lakeStates
  TSN<-merge(TSN,lakeStates,by='WBRCHCODE',all.x=T)
```

Add States to Lakes TSN table
-------------------------
* To assess changes by state we add a field for state name to TSN
* Anne Hoos assigned the MRB2 lakes to states
* Lakes also assigned to states in lakesSpatialData.Rmd (df "lakeStates" in lakesSpatialData.rda)
* State assignments compared and 34 were different
* The differences appear to be on state borders so are due to slight differences in line work
* Several were viewed in ArcGIS confirming they were on the state borders.
* Below is a list of non-matches

```{r TSNstates1, include=FALSE, echo=FALSE, cache=FALSE}
#    table(Match)   #F=34 T=23510
  #compare values for ST
    na.exclude(Compare[!Match,c(1,3,5)])
```


```{r TSNmap, include=FALSE, echo=FALSE, cache=FALSE}
    #merge coordinates data with TSN
      a<-data.frame(coordinates(lakesPtAlbers)) #get the coordinates
      a$WBRCHCODE<-lakesPtAlbers$WBRCHCODE
      TSN<-merge(TSN,a,by='WBRCHCODE',all.x=TRUE)
    #convert to spatial points data.frame
      a<-subset(TSN,!is.na(TSN$x))  #eliminate lakes without spatial data
      coords<-data.frame(x=a$x,y=a$y)
      TSNpts<-SpatialPointsDataFrame(coords,a)  #convert to spdf

#plot changes 
  plotChg<-function(est='outChg',CEX=.2){
    #plot all lakes
      plot(TSNpts,col='grey',pch=16,cex=CEX) 
    #plot lakes that have improved
      chg<-subset(TSNpts,TSNpts@data[,est]==-1 & !is.na(TSNpts$x))
      plot(chg,col='blue',pch=16,add=T,cex=CEX)
    #plot lakes that got worse (just 2)
      chg<-subset(TSNpts,TSNpts@data[,est]==1 & !is.na(TSNpts$x))
      if(length(chg)>0) plot(chg,col='red',pch=16,add=T,cex=CEX*5)
    #add title and legend
      title(est)
      if(length(chg)>0){ legend('topleft',legend=c('No Change','Better','Worse'),
                               pch=16,col=c('grey','blue','red'),bty="n")
      }else{
        legend('topleft',legend=c('No Change','Better'),
                               pch=16,col=c('grey','blue'),bty="n")
        }
  }
``` 

Results
-------------------------
* Lake trophic state was estimated  based on the inflow, outflow, and in-lake Nitrogen Concentrations for base conditions (2002) and the changes expected by 2020 under the CMAQ scenario.
* Trophic state for inflow and in-lake concentrations were calculated by the USGS for all lakes in MRB1 and MRB2
* Trophic state for outflow concentrations were done here for MRB1 but not MRB2 as we don't have the flow data (TOT_CFS) for MRB2 lakes (yet).
* Below is a table showing the number of lakes by trophic state and the number expected change based on the 2020 CMAQ Scenario

```{r table, include=TRUE, echo=FALSE, cache=FALSE}
  TSNtable
```

* Changes are seen all along the east coast.
* Below are maps for the change analysis based on:
  * outChg: outflow concentrations
  * inChg: inflow concentrations
  * lakeChg: in-lake concentrations.

```{r outChg, include=TRUE, echo=FALSE, cache=FALSE}
  #suppress warnings-if no lakes got worse this throws a warning-so warnings are turned off
      options(warn=-1)
  plotChg('outChg')
```

```{r inChg, include=TRUE, echo=FALSE, cache=FALSE}
  plotChg('inChg')
```

```{r lakeChg, include=TRUE, echo=FALSE, cache=FALSE}
  plotChg('lakeChg')
  #turn warnings back on
      options(warn=0)
```

Changes By State
-------------------------
* Below is a table that shows the changes by state
* N is the number of lakes in the state (lakes that occur in more than 1 state arbitrarily assigned)
* Better = number of lakes that improved by 1 trophic state (e.g., moved from Mesotrophic to Oligotrophic)
* NoChange = number of lakes that did not change trophic state
* Worse = number of lakes that got worse by 1 trophic state (e.g., moved from Oligotrophic to Mesotrophic)

```{r stateChg, include=TRUE, echo=FALSE, cache=FALSE}
  a<-aggregate(TSN$outChg,by=list(TSN$ST,TSN$outChg),length)
  names(a)<-c('ST','Chg','Count')
  Better<-subset(a[-2],a$Chg==-1)
    names(Better)<-c('ST','Better')
  NoChange<-subset(a[-2],a$Chg==0)
    names(NoChange)<-c('ST','NoChange')
  Worse<-subset(a[-2],a$Chg==1)
    names(Worse)<-c('ST','Worse')
  stateChg<-merge(Better,NoChange,by='ST',all.y=TRUE)
  stateChg<-merge(stateChg,Worse,by='ST',all.x=TRUE)
  stateChg[is.na(stateChg)] <- 0 #replace NA with Zeros
  stateChg$N<-apply(stateChg[2:4],1,sum)
  stateChg$BetterPercent<-round(100*stateChg$Better/stateChg$N,2)
  stateChg$WorsePercent<-round(100*stateChg$Worse/stateChg$N,2)
  stateChg<-stateChg[,c(1,5,2,3,4,6,7)]
  stateChg
```
