<<<<<<< HEAD
East Coast SPARROW: Lake Trophic State Change Analysis for the CMAQ Scenario
========================================================
Stuff to do 
-------------------------
* missing flow (TOT_CFS) data for MRB2
=======
Title
========================================================
Stuff to do 
-------------------------
* this needs to be completely reworked
* use the data from EClakes.sqlite
* Check that the names WBRCHCODE (all caps) and WBID (not WB_ID) are used consistently
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8

<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
First make sure you are in the analysis directory: 
  setwd('analysis')
  library(knitr)
<<<<<<< HEAD
  knit('trophicStateCMAQ.rmd')
=======
  knit('EClakesSQL.rmd')
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8
  
Some useful RSQLite commands
      EC<-dbConnect(SQLite(), dbname='../data/EClakes.sqlite')
      dbListTables(EC)                         # List the tables in the database
      dbListFields(EC, "R2002m1")               # List the columns in a table
      dbReadTable(EC, "R2002m2")                # Display the data in a table method1
      dbGetQuery(EC, "SELECT * from R2020m1")   # Display the data in a table method2
-->
<<<<<<< HEAD
Introduction
-------------------------
* The East Coast SPARROW model predicts nitrogen loads to streams in MRB1 and MRB2 (HUC regions 01, 02, 03, and 06)
* Information on the East Coast Model can be found here:
  * Moorman, M.C., Hoos, A.B., Bricker, S.B., Moore, R.B., García, A.M., and Ator, S.W., 2014, Nutrient load summaries for major lakes and estuaries of the Eastern United States, 2002: U.S. Geological Survey Data Series 820, 94 p.,  http://dx.doi.org/10.3133/ds820.
* Estimates of atmospheric deposition of nitrogen are derived from the CMAQ (Community Multi-scale Air Quality Model) model (http://www.epa.gov/AMD/Research/RIA/cmaq.html)
* The CMAQ model has:
  * Interpolated depostion based on 2002 data
  * Predicted deposition for 2020 following changes to air rules
* Our goal is to evaluate changes in lake trophic state between 2002 and 2020 based on the EC SPARROW model prediction for nitrogen loads to the lakes in MRB1 and MRB2

Objectives
-------------------------
* Estimate trophic state of lakes based on 2002 conditions
* Estimate trophic state of lakes based on 2020 conditions
* Analyze changes in trophic state between 2002 and 2020
* Map changes in trophic state

Data Sources
-------------------------
* The 2002 and 2020 sparrow N predictions aggregated to lakes for MRB1 and MRB2 are in: '/data/EClakes.sqlite'. 
* Data Definitions: https://github.com/willbmisled/ECsparrow/blob/master/data/EClakesSQL.md
* Includes the following tables:
  * **R2002m1**: N predictions based on 2002 conditions for lakes in MRB1
  * **R2020m1**: N predictions based on 2020 conditions for lakes in MRB1
  * **R2002m2**: N predictions based on 2002 conditions for lakes in MRB2
  * **R2020m2**: N predictions based on 2020 conditions for lakes in MRB2
* Polygon (lakes) and point (lakesPt) locations of the lakes are available in '/data/lakesSpatialData.rda'.  
* Data Definitions: https://github.com/willbmisled/ECsparrow/blob/master/data/lakesSpatialData.md
=======

>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8

```{r loadPkgs, include=FALSE, echo=FALSE, cache=FALSE}
  options(stringsAsFactors = FALSE, scipen=9) 

<<<<<<< HEAD
#function to install (if needed) and load R packages by list
=======

  #function to install (if needed) and load R packages by list
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8
    libs<-c('RSQLite','rgdal','maptools','RODBC') #list of packages to load
  
    installLoad<-function(pck)#user defined function
    {
      if(!pck%in%installed.packages()){install.packages(pck,repos="http://rweb.quant.ku.edu/cran/")}
      require(pck, character.only = TRUE)
<<<<<<< HEAD
}
=======
    }
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8
  
    lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

```{r data, include=FALSE, echo=FALSE, cache=TRUE}
#get the lakes spatial data
<<<<<<< HEAD
    load('../data/lakesSpatialData.rda')
=======
    load('./data/lakesSpatialData.rda')
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8

#transform lakesPT to albers 
  #ESRI USA Contiguous Albers Equal Area Conic (used by MRB1 WBIDLakes as AlbersX and AlbersY)
    Albers<-CRS('+proj=aea +x_0=0 +y_0=0 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +units=m +datum=NAD83')
  proj4string(lakesPt)<-proj4string(lakes)
  lakesPtAlbers<-spTransform(lakesPt,Albers)

#get the estimates of N loads under the CMAQ Scenarios
<<<<<<< HEAD
  #connect to EClakes.sqlite
      EC<-dbConnect(SQLite(), dbname='../data/EClakes.sqlite')
      dbListTables(EC) # List the tables in the database
      dbListFields(EC,"R2002m1")               # List the columns in a table
      dbListFields(EC,"R2002m2")               # List the columns in a table
  #get the 2002 estimates
    a<-dbGetQuery(EC, "SELECT WBRCHCODE,TOT_CFS,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2002m1") 
    b<-dbGetQuery(EC, "SELECT WBRCHCODE,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2002m2") #missing TOT_CFS
        b$TOT_CFS<-NA
    N2002<-rbind(a,b)
  #get the 2020 estimates
    a<-dbGetQuery(EC, "SELECT WBRCHCODE,TOT_CFS,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2020m1") 
    b<-dbGetQuery(EC, "SELECT WBRCHCODE,TN_LOAD_inflow,TN_LOAD_outflow,
                              TN_CONC_inlake,TN_CONC_inload from R2020m2") #missing TOT_CFS
        b$TOT_CFS<-NA
    N2020<-rbind(a,b)

#add outflow concentrations (for MRB1) following same methods as Ann and Rich (see ECnote.rmd)
  N2002$TN_CONC_outload<-N2002$TN_LOAD_outflow/ (N2002$TOT_CFS * 28.32 * 1/1000000 * 86400 * 365 )
  N2020$TN_CONC_outload<-N2020$TN_LOAD_outflow/ (N2020$TOT_CFS * 28.32 * 1/1000000 * 86400 * 365 )
  
#Check for missing values
  table(is.na(N2002$TN_CONC_outload))   #missing = 23681; MRB2
  table(is.na(N2020$TN_CONC_outload))   #missing = 23681; MRB2
  table(is.na(N2002$TN_CONC_inload))    #no missing values
  table(is.na(N2020$TN_CONC_inload))    #no missing values
  table(is.na(N2002$TN_CONC_inlake))    #no missing values
  table(is.na(N2020$TN_CONC_inlake))    #no missing values
  
#Check number of observations and order of reachcode
  nrow(N2002)  #41566
  nrow(N2020)  #41566
  all.equal(N2002$WBRCHCODE,N2020$WBRCHCODE)
```


```{r TSNtable, include=FALSE, echo=FALSE, cache=TRUE}
#select TS criteria
  criteriaNLA<-c(.35,.75,1.4)  #mg/l
#function to assign Trophic State in lakes based on Nitrogen thresholds
  #' criteria=thresholds between the trophic states; default=criteriaNLA
  #' conc=which of 3 estimates of lake concentration to use; default=N2002$TN_CONC_inlake
  
  estTSN<-function(criteria=criteriaNLA,conc=N2002$TN_CONC_inlake){ 
=======
  con <- odbcConnectAccess('C:/Bryan/EPA/Data/Sparrow/EastCoast/EC2014.mdb')
   L2002 <- sqlQuery(con, "SELECT mrb1_massbalancereservoirs_2002.* FROM mrb1_massbalancereservoirs_2002")
   L2020 <- sqlQuery(con, "SELECT mrb1_massbalancereservoirs_2020.* FROM mrb1_massbalancereservoirs_2020")
  close(con)
  
#add outflow concentration following same methods as Ann and Rich (see ECnote.rmd)
  L2002$TN_CONC_outload<-L2002$TN_LOAD_outflow/ (L2002$TOT_CFS * 28.32 * 1/1000000 * 86400 * 365 )
  L2020$TN_CONC_outload<-L2020$TN_LOAD_outflow/ (L2020$TOT_CFS * 28.32 * 1/1000000 * 86400 * 365 )
  
#No missing values
  table(is.na(L2002$TN_CONC_outload))
  table(is.na(L2020$TN_CONC_outload))
  table(is.na(L2002$TN_CONC_inload))
  table(is.na(L2020$TN_CONC_inload))
  table(is.na(L2002$TN_CONC_inlake))
  table(is.na(L2020$TN_CONC_inlake))
  
#data in same order? No but it can be fixed
  nrow(L2002)
  nrow(L2020)
  all.equal(L2002$WBRCHCODE,L2020$WBRCHCODE)
  
  L2002<-L2002[order(L2002$WBRCHCODE),]
  L2020<-L2020[order(L2020$WBRCHCODE),]
  all.equal(L2002$WBRCHCODE,L2020$WBRCHCODE)
```


```{r tblTSN, include=FALSE, echo=FALSE, cache=TRUE}
#select TS criteria
  criteriaNLA<-c(.35,.75,1.4)  #mg/l
#function to assign Trophic State
  estTSN<-function(criteria=criteriaNLA,conc=L2002$TN_CONC_outload){ 
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8
    TSN<-factor(rep('Oligo',length(conc)),ordered=T,levels=c("Oligo","Meso","Eu","Hyper"))
    TSN[conc>criteria[1]]<-'Meso'
    TSN[conc>criteria[2]]<-'Eu'
    TSN[conc>criteria[3]]<-'Hyper'
<<<<<<< HEAD
    TSN[is.na(conc)]<-NA
=======
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8
  return(TSN)
  }

#Estimate trophic state for each lake  
<<<<<<< HEAD
  TSN<-data.frame(WBRCHCODE=N2002$WBRCHCODE)
  TSN$out2002<-estTSN(conc=N2002$TN_CONC_outload)
  TSN$out2020<-estTSN(conc=N2020$TN_CONC_outload)
    TSN$outChg<-as.numeric(TSN$out2020)-as.numeric(TSN$out2002)
  TSN$in2002<-estTSN(conc=N2002$TN_CONC_inload)
  TSN$in2020<-estTSN(conc=N2020$TN_CONC_inload)
    TSN$inChg<-as.numeric(TSN$in2020)-as.numeric(TSN$in2002)
  TSN$lake2002<-estTSN(conc=N2002$TN_CONC_inlake)
  TSN$lake2020<-estTSN(conc=N2020$TN_CONC_inlake)
    TSN$lakeChg<-as.numeric(TSN$lake2020)-as.numeric(TSN$lake2002)
 
   
#create a table of trophic state totals for each TN estimate
  tblTSN<-function(df=TSN,x='out'){
    z<-data.frame(Year=c(2002,2020),Estimate=c(x,x))
    a<-as.data.frame(table(df[,paste(x,'2002',sep='')],useNA='ifany'))[,2]
    b<-as.data.frame(table(df[,paste(x,'2020',sep='')],useNA='ifany'))[,2]
    a<-data.frame(rbind(a,b))
    #add column for missing values 
      if(ncol(a)==4) a[,5]<-0 
    #rename columns
    names(a)<-c('Oligo','Meso','Eu','Hyper','Missing')
    #add info
      z<-cbind(z,a)
    #add column for improved lakes
      chg<-na.exclude(df[,paste(x,'Chg',sep='')])
      z[2,'Improved']<-length(chg[chg<0])
    #add column for deteriorated lakes
      chg<-na.exclude(df[,paste(x,'Chg',sep='')])
      z[2,'Worse']<-length(chg[chg>0])    
    return(z)
  } 

#create table
=======
  TSN<-data.frame(WBRCHCODE=L2002$WBRCHCODE)
  TSN$out2002<-estTSN(conc=L2002$TN_CONC_outload)
  TSN$out2020<-estTSN(conc=L2020$TN_CONC_outload)
    TSN$outChg<-as.numeric(TSN$out2020)-as.numeric(TSN$out2002)
  TSN$in2002<-estTSN(conc=L2002$TN_CONC_inload)
  TSN$in2020<-estTSN(conc=L2020$TN_CONC_inload)
    TSN$inChg<-as.numeric(TSN$in2020)-as.numeric(TSN$in2002)
  TSN$lake2002<-estTSN(conc=L2002$TN_CONC_inlake)
  TSN$lake2020<-estTSN(conc=L2020$TN_CONC_inlake)
    TSN$lakeChg<-as.numeric(TSN$lake2020)-as.numeric(TSN$lake2002)
  
  
  tblTSN<-function(df=TSN,x='out'){
    print(table(df[,paste(x,'2020',sep='')],df[,paste(x,'2002',sep='')]))
    print(table(df[,paste(x,'2002',sep='')]))
    print(table(df[,paste(x,'2020',sep='')]))
    print(table(df[,paste(x,'Chg',sep='')]))
  }  
   
   
#create a table of trophic state totals for each TN estimate
  tblTSN<-function(df=TSN,x='out'){
    z<-data.frame(Year=c(2002,2020),estimate=c(x,x))
    a<-as.data.frame(table(df[,paste(x,'2002',sep='')]))[,2]
    b<-as.data.frame(table(df[,paste(x,'2020',sep='')]))[,2]
    a<-rbind(a,b)
    z<-cbind(z,a)
    z[2,'improved']<-as.data.frame(table(df[,paste(x,'Chg',sep='')]))[1,2]
    names(z)[3:6]<-c('Oligo','Meso','Eu','Hyper')
    return(z)
  }  
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8
   a<-rbind(tblTSN(x='out'),
   tblTSN(x='in'),
   tblTSN(x='lake'))
   row.names(a)<-1:6
<<<<<<< HEAD
  TSNtable<-a
  #TSNtable
```

```{r TSNmap, include=FALSE, echo=FALSE, cache=TRUE}
    #merge coordinates data with TSN
      a<-data.frame(coordinates(lakesPtAlbers)) #get the coordinates
      a$WBRCHCODE<-lakesPtAlbers$WBRCHCODE
      TSN<-merge(TSN,a,by='WBRCHCODE',all.x=TRUE)
    #convert to spatial points data.frame
      a<-subset(TSN,!is.na(TSN$x))  #eliminate lakes without spatial data
      coords<-data.frame(x=a$x,y=a$y)
      TSNpts<-SpatialPointsDataFrame(coords,a)  #convert to spdf

#plot changes 
  plotChg<-function(est='outChg',CEX=.2){
    #plot all lakes
      plot(TSNpts,col='grey',pch=16,cex=CEX) 
    #plot lakes that have improved
      chg<-subset(TSNpts,TSNpts@data[,est]==-1 & !is.na(TSNpts$x))
      plot(chg,col='blue',pch=16,add=T,cex=CEX)
    #plot lakes that got worse (just 2)
      chg<-subset(TSNpts,TSNpts@data[,est]==1 & !is.na(TSNpts$x))
      if(length(chg)>0) plot(chg,col='red',pch=16,add=T,cex=CEX*5)
    #add title and legend
      title(est)
      if(length(chg)>0){ legend('topleft',legend=c('No Change','Better','Worse'),
                               pch=16,col=c('grey','blue','red'),bty="n")
      }else{
        legend('topleft',legend=c('No Change','Better'),
                               pch=16,col=c('grey','blue'),bty="n")
        }
  }
``` 

Results
-------------------------
* Lake trophic state was estimated  based on the inflow, outflow, and in-lake Nitrogen Concentrations for base conditions (2002) and the changes expected by 2020 under the CMAQ scenario.
* Trophic state for inflow and in-lake concentrations were calculated by the USGS for all lakes in MRB1 and MRB2
* Trophic state for outflow concentrations were done here for MRB1 but not MRB2 as we don't have the flow data (TOT_CFS) for MRB2 lakes (yet).
* Below is a table showing the number of lakes by trophic state and the number expected change based on the 2020 CMAQ Scenario

```{r table, include=TRUE, echo=FALSE, cache=TRUE}
  TSNtable
```

* Changes are seen all along the east coast.
* Below are maps for the change analysis based on:
  * outChg: outflow concentrations
  * inChg: inflow concentrations
  * lakeChg: in-lake concentrations.

```{r outChg, include=TRUE, echo=FALSE, cache=FALSE}
  #suppress warnings-if no lakes got worse this throws a warning-so warnings are turned off
      options(warn=-1)
  plotChg('outChg')
```

```{r inChg, include=TRUE, echo=FALSE, cache=FALSE}
  plotChg('inChg')
```

```{r lakeChg, include=TRUE, echo=FALSE, cache=FALSE}
  plotChg('lakeChg')
  #turn warnings back on
      options(warn=0)
```



=======
  tblTSN<-a
  tblTSN
```


#merge coordinates data with TSN
  a<-data.frame(coordinates(lakesPtAlbers)) #get the coordinates
  a$WBRCHCODE<-as.numeric(row.names(a))
  TSN<-merge(TSN,a,by='WBRCHCODE',all.x=TRUE)
#convert to spatial points data.frame
  a<-na.exclude(TSN)
  coords<-data.frame(x=a$x,y=a$y)
  TSNpts<-SpatialPointsDataFrame(coords,a)  #convert to spdf

#plot changes  
  plotChg<-function(est='outChg',CEX=.2){
    plot(TSNpts,col='grey',pch=16,cex=CEX)
    plot(TSNpts[TSNpts@data[,est]==-1,],col='blue',pch=16,add=T,cex=CEX)
    title(est)
  }

windows(10.5,7)
par(mfrow=c(1,3))
  plotChg('outChg')
  plotChg('inChg')
  plotChg('lakeChg')
  
>>>>>>> e265630f21b75e693fc6215f87103ad8ec78fed8
  
